---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Wisneski, Kelly"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

-----

Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)
library(RColorBrewer)
my_palette <- brewer.pal(name="GnBu", 9)[4:9]

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",")
mydata$CLASS <- factor(mydata$CLASS)
mydata$SEX <- factor(mydata$SEX)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."
mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use table() to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}

summary(mydata)
table(mydata$CLASS, mydata$RINGS)
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: Variables SEX and CLASS are both categorical or qualitative variables. All other variables are numeric or quantitative variables. HEIGHT appears to be the variable closest to a symmetrical normal distribution, with mean and median fairly similar and a skewness of -0.225262. LENGTH (skewness=-.67) and DIAM (skewness=-0.62) are both negatively skewed, which suggest the presence of low outliers causing the mean to be lower than the median. WHOLE (skewness=.047), SHUCK (skewness=0.64), RINGS (skewness=1.24), VOLUME (skewness=.44), and RATIO (skewness=.71) are all positively skewed, which suggests the presence of high outliers causing the mean to be greater than the median.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
t <- table(mydata$SEX, mydata$CLASS)
addmargins(t)

freq <- data.frame(t)
colnames(freq) <- c('Sex', 'Class', 'Frequency')
ggplot(freq, aes(x=Class, y=Frequency, fill=Sex, group=Sex)) + 
  geom_bar(stat="identity", position=position_dodge()) +
  ggtitle("Abalone by Class | Sex") +
  scale_fill_manual(values=my_palette)
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: It is surprising that there is a good number of infants in both A4 and A5, which are supposed to be the oldest classes. It is likely that these abalones could not be classified as male or female. Additionally, there are more infants in A2 than in A1, which suggests that counting rings is an inaccurate or difficult method for determining age. It is also surprising the that frequency of each sex varies so greatly by class. Is there an undetermined third factor that causes there to be a great number of males in category A3, but drop significantly by A5? I would expect male/female numbers to increase gradually over time as more infants are maturing and being properly sexed. I wonder if this is another classification issue skewing the data?***

(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
srs <- sample(seq(1:1036), 200)
work <- mydata[srs,]
plot(work[, 2:6], cex=.5, lwd=.3)

```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}
ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
  geom_point(aes(color = CLASS),size = 1) + 
  ggtitle(bquote('Whole weight (grams) | Volume ('~cm^3~')')) +
  scale_color_brewer(palette="GnBu") +
  ylab("Whole weight (grams)") +
  xlab(bquote('Volume ('~cm^3~')'))
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + 
  geom_point(aes(color = CLASS),size = 1,) +
  scale_color_brewer(palette="GnBu") +
  ggtitle('Shuck weight (grams) | Whole weight (grams)') +
  geom_abline(intercept=0, slope=max(mydata$SHUCK/mydata$WHOLE), size=.25, linetype=2) +
  ylab("Shuck weight (grams)") +
  xlab('Whole weight (grams)') 
```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: The first plot appears more variable than the second plot. This indicates that there is a stronger correlation between the two weights (shuck versus whole) than between volume and weight. Variability appears consistent across classes in the first plot. In the second plot, however, there appears to be a kind of "banding" by class: the older the abalone, the more variable the data appears, trending towards a lower shuck-to-weight ratio. This might indicate that as abalone mature, their shell might be growing heavier than then their meat (the portion that is "shucked").***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a, fig.height = 7, fig.width = 7}
h <- ggplot(mydata, aes(RATIO, fill=SEX)) +
  facet_grid(. ~ SEX, space="free") +
  geom_histogram(bins=14, color="black", size=.2) + 
  scale_fill_manual(values=my_palette) +
  ggtitle("Histograms of Ratio") +
  theme(legend.position="none", axis.title.x=element_blank()) +
  ylab("Frequency")
  
b <- ggplot(mydata, aes(y=RATIO, fill=SEX)) +
  facet_grid(cols = vars(SEX)) +
  geom_boxplot(color="black", size=.2, outlier.size=2, outlier.shape=1) + 
  scale_fill_manual(values=my_palette) +
  ggtitle("Boxplots of Ratio") +
  theme(legend.position="none", axis.title=element_blank())

q <- ggplot(mydata, aes(sample=RATIO, color=SEX)) +
  facet_grid(cols = vars(SEX)) +
  geom_qq(cex=2, shape=1) + 
  scale_color_manual(values=my_palette) +
  ggtitle("QQ Plots of Ratio") +
  geom_qq_line(color="black", cex=.5) +
  theme(legend.position="none") +
  ylab("Sample Quantities") +
  xlab("Theoretical Quantities")

grid.arrange(h, b, q, nrow=3)
```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: From the displays, we see that all distributions are skewed to the right, and appear at first glance to be non-normally distributed. This is likely due to the mild and extreme outliers in all three graphs, but in particular in the female and infant distributions. All three plots for each graphs show a positive skew: the histograms display long right tails, the boxplots have IQRs on the low end with outliers on the upper end, and the QQ plot values trend above the QQ line, especially for the most positive values. While the female distribution appears to have the most extreme outliers, the infant distribution appears to have the greatest quantity of outliers.***

(3)(b) (2 points) Use the boxplots to identify RATIO outliers (mild and extreme both) for each sex. Present the abalones with these outlying RATIO values along with their associated variables in "mydata" (Hint:  display the observations by passing a data frame to the kable() function).  

```{r Part_3b}
i_out <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "I"])$out, ]
f_out <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "F"])$out, ]
m_out <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "M"])$out, ]
kable(rbind(i_out, f_out, m_out))
```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: Most shuck/volume ratio outliers come from the infant classification. Further, the large majority of outliers are from the younger classes, A1-A3; only one outlier is classified as A5. Additionally, the only extreme outliers come from the female and infant classes, but none appear for the male class. The number of outliers in these younger classes might be skewing the data and make drawing correlation between physical measurements and age difficult.***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
b_v <- ggplot(mydata, aes(x=CLASS, y=VOLUME, fill=CLASS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  scale_fill_brewer(palette="GnBu") +
  ggtitle("Boxplots of Volume | Class") +
  theme(legend.position="none")

b_w <- ggplot(mydata, aes(x=CLASS, y=WHOLE, fill=CLASS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  scale_fill_brewer(palette="GnBu") +
  ggtitle("Boxplots of Whole weight | Class") +
  theme(legend.position="none")

s_v <- ggplot(data = mydata, aes(x = RINGS, y = VOLUME)) + 
  geom_point(size = .5,) +
  scale_color_brewer(palette="GnBu") +
  ggtitle("Volume | Rings") +
  theme(legend.position="none")

s_w <- ggplot(data = mydata, aes(x = RINGS, y = WHOLE)) + 
  geom_point(size = .5,) +
  scale_color_brewer(palette="GnBu") +
  ggtitle("Whole weight | Rings") +
  theme(legend.position="none")

grid.arrange(b_v, b_w, s_v, s_w, nrow=2)

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: Neither of these variables perform well as predictors of age. Both volume and whole weight are positively correlated to the number of rings -- that is, as either weight or volume increases, so does the number of rings. However, this is a very loose correlation. For example, an abalone with 500 cm^3 volume (abalone with 150g whole weight) could have anywhere between 8 rings and 23 rings. The story is similar for the correlation between volume & class and weight & class: generally higher volume or higher weight abalone are classified higher/older, but especially classes A3, A4, and A5 have strikingly similar IQRs for both volume and weight. An abalone of volume 250 cm^3 or weight 125g could very likely be in either A3, A4, or A5. This either indicates that volume and weight are poor predictors of exact age (though good predictors of a ballpark age), or that the classification system for age is flawed. This aligns with the knowledge that ring counting is a difficult and laborious process.***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
agg <- aggregate(mydata[c('VOLUME', 'SHUCK', 'RATIO')], FUN=mean, by=list(mydata$SEX, mydata$CLASS))

m_v <- matrix(agg$VOLUME, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

m_s <- matrix(agg$SHUCK, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

m_r <- matrix(agg$RATIO, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

kable(m_v, caption="Volume")
kable(m_s, caption="Shuck")
kable(m_r, caption="Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.height = 10, fig.width = 6}
colnames(agg) <- c('Sex', 'Class', 'Volume', 'Shuck', 'Ratio')
c_r <- ggplot(data = agg, aes(x = Class, y = Ratio, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Ratio | Class") +
  scale_color_manual(values=my_palette)

v_r <- ggplot(data = agg, aes(x = Class, y = Volume, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Volume | Class") +
  scale_color_manual(values=my_palette)

s_r <- ggplot(data = agg, aes(x = Class, y = Shuck, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Shuck weight | Class") +
  scale_color_manual(values=my_palette)

grid.arrange(c_r, v_r, s_r, nrow=3)
```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: Generally, for all sexes, we see mean shuck/volume ratio decreasing with age (class) and both volume and shuck increasing with age. However, these plots raise a number of questions about the variations in shuck and volume between the genders. Namely, why do females trend heavier and larger than both males and infants? Further, why do females trend heavier/larger but their growth over time is smaller than males and infants? Does the small difference between female A1 and A2 (volume and shuck) indicate that females develop more slowly? Why are infants consistently both lighter and smaller than the adults, but also have a lower shuck/volume ratio? Is this due to infants truly having a lower shuck/volume ratio, or is just more difficult to identify the sex of an abalone with a lower shuck/volume ratio? Why do both adults and infants tend to shrink (get lighter/smaller) between A4 and A5?***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
i_v <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="lightseagreen") + 
  ggtitle("Infant Volume | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 1000)) +
  ylab("Volume") +
  xlab("Rings")

a_v <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="steelblue2") + 
  ggtitle("Adult Volume | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 1000)) +
  ylab("Volume") +
  xlab("Rings")
  
i_w <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="lightseagreen") + 
  ggtitle("Infant Whole weight | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")

a_w <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="steelblue2") + 
  ggtitle("Adult Whole weight | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")

grid.arrange(i_v, a_v, i_w, a_w, nrow=2)
```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: These displays suggest that most abalone growth occurs in the first half of the abalone life span, from 1 ring to 10-11 rings; the volume and weight weight both increase most rapidly during this period. By contrast, growth appears to level, or even shrink, after reaching 10-11 rings. Additionally, while it makes sense that infants tend to have lower volumes and weights than adults, there is a surprising amount of overlap between the two displays. For example, if classifying on volume/weight alone, an abalone of volume 500 cm^3 or 150g is pretty likely an adult, but any volume or weight lower than this threshold could easily be either infant or adult. Most abalone classified as "infants" have the same size and weight of some of the smaller/lighter adults. Lastly, the infants tend to be closer to volume/weight to each other; except for a group of highly variable 12-ring infants, the infant displays have smaller IQRs and smaller standard deviations as compared to the adults, which are more variable in volume and weight within each ring grouping overall.***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: It is likely the original study was not successful because the erroneous classification of infants was not taken into account. Because immature abalone are difficult to correctly sex, it might be difficult to determine what is an immature abalone, and what is simply a small adult abalone. This sex information might be important to take into account when trying to predict age from physical measurements. Additionally, the growth of abalone (by size and by weight) seems to taper after the abalone have 10-11 rings; weight and volume do not increase proportionally with increase in rings after about 10-11 rings. In young abalone, when low weight/volume have a stronger correlation to fewer rings, the sex of the abalone can a big factor in the size/weight (females tend to start heavier/bigger than males and infants as seen in the 5b displays). In the later years, as seen in the 5c displays, the size/weight of an abalone does not increase proportionally with the number of rings; compounded with the fact that there is not a strong correlation between volume-rings and weight-rings to begin with (4b displays), it is therefore nearly impossible to tell the age of any A3-A5 abalone given physical measurements alone. In conclusion, inaccurate sex classifications make it difficult to predict age from physical measurements for young abalone, and a slow in weight/volume growth in A3-A5 make the prediction difficult for older abalone.***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: I would want to know more about the sample and population. For example, what sampling methods were used? Was a simple random sample used, or was it a non-random sampling method? How was the sampling frame constructed? When was the sample taken (is the data outdated)? What is the sample size, and how does it compare to the population size? What population parameters are known, or is it possible to estimate them? Are there known biases or other known influential factors that might affect how representative the sample is of the population? There are a great number of factors that could determine whether or not the sample is representative of the sampled population.***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: It is difficult to analyze data derived from observation studies because there are a great number of factors at play. Human error and bias -- whether from the construction of the study, collection of data, or analysis of result -- can drastically affect observational studies. It is easy for an observational study to miss or overlook any number of explanatory variables that could contribute to the response variable. Without a control group, correlation is the best that an observational study can hope for. It is not possible to determine causality from an observational study, but an observational study can help determine variables of interest that could potentially be studied further to determine causality.***
